An introduction to Docker for reproducible research

### 2014-11-04 docker-paper/manuscript.Rmd
ticking off more edits

A recent study by researchers at the University of Arizona found that
less than 50% of software could even be successfully built or installed
[@Collberg_2014] and similar results are seen in an ongoing effort by other researchers to
replicate that study [@Krishnamurthi_2014]. Installing or building software
necessary to run the code in question assumes the ability to recreate
the computational environment of the original researchers. 
<!-- Time savings. Personal reproducibility. Collaboration. Teaching. --> 
<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents
-->
<!-- Omit this? -->

@@

Docker is an open source project that builds on many long-familiar
technologies from operating systems research: LXC containers,
virtualization of the OS, and a hash-based or git-like versioning
and differencing system, among others.

@@

<!-- FIXME move this later --> 
To simplify this process, Docker also supports *Automated Builds*
through the Docker Hub ([hub.docker.com](http://hub.docker.com)).  This acts as a kind of Continuous Integration
(CI) service that verifies the image builds correctly whenever the
Dockerfile is updated, particularly if the Dockerfile includes checks
for the environment. The Hub also provides a convenient distribution
service, freely storing the pre-built images, along with their metadata,
for download and reuse by others. The Docker Hub is a free service and
an open source software product so that users can run their own private
versions of the Hub on their own servers, for instance, if security of
the data or the longevity of the public platform is a concern.

@@

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [_e.g._ @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces like RStudio
Server are rapidly filling this gap).

@@

@@

Sharing these images is further facilitated by the Docker Hub technology. 
While Docker images tend to be much smaller than equivalent virtual
machines, moving around even 100's of gigabytes can be a challenge.
In their work on virtual machines, Dudley _et al._ [@Dudley_2010] recommend only running
these tools on cloud servers such as the Amazon EC2 system. As most
researchers still develop software locally and may not have ready access
to these resources, such a requirement adds an additional barrier to
reuse.  Docker provides a convenient way to share any image publicly
or privately through the Hub after creating a free account.  Docker Hub
is both a publicly available service and also available as 
a separate open source platform that can be deployed on private servers.

@@

The approach of Linux Containers represented by Docker offers a technical
solution to what is frequently seen as the primary weakness of the
standard virtual machine approach to reproducibility - reusing and
remixing elements.  To some extent this is already addressed by the
DevOps approach of Dockerfiles, providing a scripted description
of the environment that can be tweaked and altered, but also includes
something much more fundamental to Docker.

@@

```
FROM ubuntu:latest
RUN apt-get update 
RUN apt-get -y install r-recommended
```

This acts like a software dependency; but unlike other software, a
Dockerfile must have exactly one dependency (one `FROM` line). Note that
a particular version of the dependency can be specified using the `:` notation.
One can in fact be much more precise, declaring not only version numbers
like `ubuntu:14:04`, but specific cryptographic hashes that ensure we
get the exact same image every time.

One can now build this Dockerfile and give it name, by running in the
working directory:

```bash
Docker build -t username/r-recommended .
```

@@

If this image is shared (either directly or through a Docker Hub),
another user can now build directly on our image, rather than on the
`ubuntu:latest` image used as a base in the first Dockerfile.  To do
so, one simply specifies a different image in the `FROM` directive of
the Dockerfile, followed by the lines required to add any additional
software required.

@@

```bash
docker run -d --name db training/postgres
docker run -d -P --link db:db training/webapp python app.py
```

Unlike the much more heavyweight virtual machine approach, containers
are implemented in way such that a single computer can easily run 100s
of such services each in their own container, making it easy to break
computational elements down into logically reusable chunks that come,
batteries included, with everything they need to run reproducibly.
A researcher could connect a container providing a computational 
environment for a different language to this same PostgreSQL container, 
and so forth.

@@

In addition to version managing the Dockerfile, the images themselves are
versioned using a git-like hash system (_e.g._ see `Docker commit`, `docker
push`/`Docker pull`, `docker history`, `docker diff`).  Docker images and
containers have dedicated metadata specifying the date, author, parent
image, and other details (see `Docker inspect`). One can roll back an
image through the layers of history of its construction, then build off
an earlier layer, or roll back changes made interactively in a
container. For instance, here I inspect recent changes made to the
`ubuntu:latest` image:

```bash
docker history ubuntu:latest
```

@@

```bash
docker tag 25f ubuntu:latest
```
If one now inspects the history, which shows that it now begins from this earlier point:

```bash
docker history ubuntu:latest
```

@@

- _Archive tarball snapshots_. Despite similar semantics to git, Docker's
versioning system works rather differently than version management
of code.  Docker can roll back AUFS layers that have been added to
an image, but not revert to the earlier state of a particular layer.
In consequence, to preserve a bitwise identical snapshot of a container
used to generate a given set of results, it is necessary to archive the
image tarball itself -- one can not simply rely on the Docker history
to recover an earlier state.

>>>>

A recent study by researchers at the University of Arizona found that
less than 50% of software could even be successfully built or installed
[@Collberg_2014]. Though sufficiently knowledgeable users may be able
to overcome the issues in at least some of these cases [^6], similar
results are seen in an ongoing effort by other researchers to replicate
that study [@Krishnamurthi_2014], and has also been observed in other
indepedent studies [@Lapp_2014; @Garijo_2013].  Installing or building
software necessary to run the code in question assumes the ability to
recreate the computational environment of the original researchers.

[^6]: See discussion at https://gist.github.com/pbailis/9629050 and https://gist.github.com/samth/9641364

@@

Docker is an open source project that builds on many long-familiar
technologies from operating systems research: LXC containers,
virtualization of the OS, and a hash-based or git-like versioning
and differencing system, among others (see [docs.docker.com/faq](https://docs.docker.com/faq/#what-does-docker-add-to-just-plain-lxc) for
an excellent overview of what Docker adds to plain LXC).

@@

@@

<!-- Part of the reason is better image distribution-->
While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [_e.g._ @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.

rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces like RStudio
Server are rapidly filling this gap).

@@

<!--

@@

Sharing these images is further facilitated by the Docker Hub technology
([hub.docker.com](http://hub.docker.com)).  While Docker images tend
to be much smaller than equivalent virtual machines, moving around
even 100's of gigabytes can be a challenge.  The Docker Hub provides a
convenient distribution service, freely storing the pre-built images,
along with their metadata, for download and reuse by others. The Docker
Hub is a free service and an open source software product so that users
can run their own private versions of the Hub on their own servers, for
instance, if security of the data or the longevity of the public platform
is a concern.  Docker also supports *Automated Builds* through the Docker
Hub.  This acts as a kind of Continuous Integration (CI) service that
verifies the image builds correctly whenever the Dockerfile is updated,
particularly if the Dockerfile includes checks for the environment.

@@

The approach of Docker offers a technical
solution to what is frequently seen as the primary weakness of the
standard virtual machine approach to reproducibility - reusing and
remixing elements.  To some extent this is already addressed by the
DevOps approach of Dockerfiles, providing a scripted description
of the environment that can be tweaked and altered, but also includes
something much more fundamental to Docker.

@@

```
FROM ubuntu:14.04
RUN apt-get update 
RUN apt-get -y install r-recommended
```

This acts like a software dependency; but unlike other software, a
Dockerfile must have exactly one dependency (one `FROM` line). Note
that a particular version of the dependency can be specified using the
`:` notation (similarly, package managers like `apt` show here can be
directed to install specific versions, if desired. One can in fact be much
more precise by specific cryptographic hash of a particular binary image.
In practice, using hashes in FROM directives may be undesirable as it also
avoids adopting security updates that may have been made to the base image.

One can name and build this Dockerfile by running in the
working directory:

```bash
docker build -t username/r-recommended .
```

@@

If this image is shared (either directly or through a Docker Hub),
another user can now build directly on our image, rather than on the
`ubuntu:14.04` image used as a base in the first Dockerfile.  To do
so, one simply specifies a different image in the `FROM` directive of
the Dockerfile, followed by the lines required to add any additional
software required.

@@

```bash
docker run -d --name db training/postgres
docker run -d -P --link db:db training/webapp python app.py
```
This separates the task fo running the database (first line) from
the application used to analyze the data (second line), allowing a more
modular approach to reuse: another researcher could use the same database
container while connecting it to different scripts.

Unlike the much more heavyweight virtual machine approach, containers
are implemented in way such that a single computer can easily run 100s
of such services each in their own container, making it easy to break
computational elements down into logically reusable chunks that come,
batteries included, with everything they need to run reproducibly.

@@

In addition to version managing the Dockerfile, the images themselves are
versioned using a git-like hash system (_e.g._ see `Docker commit`, `docker
push`/`Docker pull`, `docker history`, `docker diff`).  Docker images and
containers have dedicated metadata specifying the date, author, parent
image, and other details (see `Docker inspect`). One can roll back an
image through the layers of history of its construction, then build off
an earlier layer, or roll back changes made interactively in a
container. For instance, here I inspect recent changes made to the
`ubuntu:14.04` image:

```bash
docker history ubuntu:14.04
```

@@

```bash
docker tag 25f ubuntu:14.04
```
If one now inspects the history, which shows that it now begins from this earlier point:

```bash
docker history ubuntu:14.04
```

@@

- _Archive tarball snapshots_. Despite similar semantics to git, Docker's
versioning system works rather differently than version management
of code.  Docker can roll back layers[^5] that have been added to
an image, but not revert to the earlier state of a particular layer.
In consequence, to preserve a bitwise identical snapshot of a container
used to generate a given set of results, it is necessary to archive the
image tarball itself -- one can not simply rely on the Docker history
to recover an earlier state.

[^5]: Technically AUFS (advanced multi layered unification filesystem )
layers, see [wikipedia.org/wiki/aufs](http://en.wikipedia.org/wiki/aufs)
